Value-Suppressing Uncertainty Maps

### 2017-09-14 paper/main.tex
Revisions / comments for related work.

There are many methods for visualizing uncertainty. MacEachren et al. ~\cite{maceachren1992visualizing} presents a theoretical survey of uncertainty visualization generally. Relevant to our domain of heatmaps and thematic maps is work in GIS and cartography; for a survey of uncertainty visualization techniques in GIS, see Kinkeldey et al. \cite{kinkeldey2014assess,kinkeldey2017evaluating}. Yet, despite these wide variety of methods, to our knowledge, there are no established standards for visualizing uncertainty in heatmaps.

@@

One challenge to conveying uncertainty information is that the decision to explicitly encode uncertainty increases the dimensionality of the data, and so requires the use of (at least one) additional visual channel ~\cite{brodlie2012review}. Uncertainty therefore inherently increases the visual complexity of a visualization. When the data are already complex to convey, and many of the more common or accurate visual variables are in use, allocating an additional dimension is non-trivial. As the number of dimensions increases, finding visual variables that are both perceptually accurate (in either estimation of quantity or discrimination of category) as well as perceptually separable from all the other encoding channels, becomes more and more difficult.

A further hurdle is that not all visual variables are well-suited for conveying uncertainty. MacEachren \ea~\cite{maceachren2012visual} evaluate a number of visual variables with respect to their \emph{semiotic fit} for representing uncertainty. They observe that certain visual variables such as blurriness and transparency seem to have a more intuitive connection to uncertainty than other variables such as shape or hue. Unfortunately, Boukhelifa \ea~\cite{boukhelifa2012evaluating} find that many visual variables habitually used for conveying uncertainty, such as blur and value, are also difficult to estimate. This results in a preference/performance gap where designers must choose between encoding uncertainty in a way that is intuitive but error prone, or use higher fidelity channels that may be more difficult to interpret.

@@

For spatial visualizations like choropleth maps, heatmaps, and treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is often encoded using color. Colors in such univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care based on the data types and important features to be visualized~\cite{rogowitz1996not}.

A design consideration when constructing color maps is whether to quantize data into a discrete set of colors, or to encode data using a continuous mapping from value to color. While continuous color maps afford greater fidelity in presenting values~\cite{muller1979perception}, non-linearity in human color perception introduces errors in extracting numeric values from continuous colors~\cite{borland2007rainbow}. Quantizing a color map is therefore an exercise in balancing \emph{perceptual} error and \emph{quantization} error~\cite{dobson1973choropleth}. Discrete maps offer a finer control over this balance, which can result in better performance in tasks involving heatmaps~\cite{padilla2017evaluating}. As such, we limit our discussion to color maps with discrete sets of colors.

@@

While uncertainty can be visualized entirely separately from the rest of the data (for instance, in a juxtaposed chart), this runs the risk that uncertainty information becomes \emph{ignorable}\cite{moritz2017trust}. It also complicates analysis, as interpreting the value and uncertainty of a given data point requires consulting two separate charts. For our goal on integrating uncertainty information into analysis, we focus instead on the construction of \emph{bivariate maps}, where value and uncertainty information is displayed simultaneously in the same visualization.

Bivariate maps can, in principle, be constructed from the combination of any two visual variables (such as shape and color, or size and texture). E.g., Ware's ``textons'' overlay glyphs on top of regions to simultaneously encode two quantitative values~\cite{ware2009quantitative}. However, in this work, our focus is on heatmaps and other sorts of thematic maps. In visualizations like heatmaps, common visual channels such as position and size are already in use to encode other information (such as spatial location and extent), and color is reserved for value. Bivariate maps in these settings therefore typically rely on a visual channel related to color for encoding a secondary variable, such as alpha~\cite{roth2010value} or pixel noise~\cite{lucchesi2017visualizing}.

Due to the perceptual integrality of visual channels~\cite{garner1970integrality, ware2012information}, the construction of color-based bivariate maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 4x4 matrix as in \figref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. For VSUPs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analytics tasks, rather than uniformly.

In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret \cite{wainer1980empirical}.

>>>>

There are many methods for visualizing uncertainty. MacEachren \ea~\cite{maceachren1992visualizing} present a theoretical survey of uncertainty visualization. Relevant to our domain of heatmaps and thematic maps \textbf{JH: where did this come from? There is no mention of focusing on this domain in the intro. The whole carto/GIS stuff seems to come out of left field.} is work in GIS and cartography; for a survey of uncertainty visualization techniques in GIS, see Kinkeldey \ea~\cite{kinkeldey2014assess,kinkeldey2017evaluating}. Yet, despite a wide variety of methods, to our knowledge, there are no established standards for visualizing uncertainty in heatmaps.

@@

One challenge of uncertainty visualization is that the decision to explicitly encode uncertainty increases the dimensionality of the data, and so requires the use of (at least one) additional visual channel~\cite{brodlie2012review}. Uncertainty therefore inherently increases the visual complexity of a visualization. When the data are already complex to convey, and many of the more common or accurate visual variables are in use, allocating an additional dimension is non-trivial. As the number of dimensions increases, finding visual variables that are both perceptually accurate (in either estimation of quantity or discrimination of category) as well as perceptually separable from all the other encoding channels, becomes increasingly difficult.

A further hurdle is that not all visual variables are well-suited for conveying uncertainty. MacEachren \ea~\cite{maceachren2012visual} evaluate a number of visual variables with respect to their \emph{semiotic fit} for representing uncertainty. They observe that certain visual variables such as blurriness and transparency seem to have a more intuitive connection to uncertainty than other variables such as shape or hue. Unfortunately, Boukhelifa \ea~\cite{boukhelifa2012evaluating} find that many visual variables habitually used for conveying uncertainty, such as blur and value, are also difficult to estimate. This results in a preference/performance gap where designers must choose between encoding uncertainty in a way that is intuitive but error prone, or use higher fidelity channels that may not intuitively convey uncertainty.

@@

For spatial visualizations like choropleth maps, heatmaps, and treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is often encoded using color. Colors in univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care~\cite{rogowitz1996not}.

A recurring design consideration when constructing color maps is whether to quantize data into a discrete set of colors, or to encode data using a continuous mapping. While continuous color maps afford greater fidelity in presenting values~\cite{muller1979perception}, non-linearity in human color perception introduces errors in extracting numeric values from continuous colors~\cite{borland2007rainbow}. Quantizing a color map is therefore an exercise in balancing \emph{perceptual} error and \emph{quantization} error~\cite{dobson1973choropleth}. Discrete maps offer finer control over this balance, which can result in better performance in tasks involving heatmaps~\cite{padilla2017evaluating}. As such, we limit our discussion to color maps with discrete sets of colors.

@@

While uncertainty can be visualized entirely separately from the rest of the data (for instance, in a juxtaposed chart), this runs the risk that uncertainty information becomes \emph{ignorable}~\cite{moritz2017trust}. It also complicates analysis, as interpreting the value and uncertainty of a given data point requires consulting two separate charts. To integrate uncertainty information into an analysis, we focus on the construction of \emph{bivariate maps}, where value and uncertainty information is displayed simultaneously in a spatially co-located manner.

Bivariate maps can, in principle, be constructed from the combination of any two visual variables (such as shape and color, or size and texture). For example Ware's ``textons''~\cite{ware2009quantitative} overlay glyphs on top of regions to simultaneously encode two quantitative values. However, in this work, our focus is on heatmaps and other sorts of thematic maps. In visualizations like heatmaps, common visual channels such as position and size are already in use to encode other information (such as spatial location and extent), and color is reserved for value \textbf{JH: this is a bit redundant with the previous subsection on color maps. It feels a bit artificial having these two sections separate. Color is a subset of bivariate map options... and we start talking about it again in the next paragraph.}. Bivariate maps in these settings therefore typically rely on a visual channel related to color for encoding a secondary variable, such as opacity (alpha)~\cite{roth2010value} or pixel noise~\cite{lucchesi2017visualizing}.

Due to the perceptual integrality of visual channels~\cite{garner1970integrality, ware2012information}, \textbf{JH: We mention integrality but never give a brief definition or explanation. I wouldn't assume that all CHI readers are familiar with this.} the construction of color-based bivariate maps is difficult. Viewers need to be able to independently assess the variables of interest \textbf{JH: They do? I thought the whole point of VSUPs is to confound indepenent assessment of value!}, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 4x4 matrix as in \figref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. For VSUPs, we adapt an insight from Dunn~\cite{dunn1989dynamic} that these limited categories ought to be assigned with regards to their importance to analysis tasks, rather than uniformly.

In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret~\cite{wainer1980empirical} \textbf{JH: It seems there is a lot ``hiding'' behind this citation. In what ways is it hard? In what ways do people fail? Any more flavor here, even if very brief? And might this motivate something akin to VSUPs?}.

\textbf{JH: This would be a natural place to put some connective prose that transitions to VSUPs...}
